The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-12 manuscript/manuscript.Rnw
use citet and p{} in the gestational period table
discuss CYP19A1 and HSDB1 expression levels

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 135 & 4 & \DLA{Cite Amazing Armadillo: geography book}\\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

@@

Reproductive hormones are central to the regulation of pregnancy and
their source, timing, and quantity can vary widely from species to
species \cite{Carter2012:Evolutionofplacentalfunction}. Expression and synthesis of hormones is
perhaps the most well-documented physiological process in the placenta
in many species. Here we quantified the extreme disparity in genes
with hormone products or enzymatic products that directly produce
hormones in multiple mammalian species (\DLA{Table 8}).

In both ruminants and primates, the placenta is responsible for the
production of estrogen and progesterone. High relative expression of
CYP19A1, the enzyme that synthesizes estrogen, and HSD3B1, the enzyme
that synthesizes progesterone, in the human, bonobo, and cow,
generally matches the high levels of placental production of estrogen
and progesterone in humans, other primates, and ruminants in late
pregnancy. In addition, HSD3B1 shows high levels in dog, which matches
the high levels of placental production of progesterone at term in
that species. \DLA{I think we need citations for this}

>>>>

\begin{table}
  \begin{tabular}{c c c p{3in}}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 135 & 4 & \DLA{Cite Amazing Armadillo: geography book}\\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

@@

Reproductive hormones are central to the regulation of pregnancy and
their source, timing, and quantity can vary widely from species to
species \cite{Carter2012:Evolutionofplacentalfunction}. Expression and synthesis of hormones is
perhaps the most well-documented physiological process in the placenta
in many species. Here we quantified the extreme disparity in genes
with hormone products or enzymatic products that directly produce
hormones in multiple mammalian species (\cref{fig:hormone_genes}).

In both ruminants and primates, the placenta is responsible for the
production of estrogen and progesterone. High relative expression of
\textit{CYP19A1} (\cref{fig:estrogen_evolution_tree}), the enzyme that
synthesizes estrogen from androstenedione is seen in \human, \bonobo,
\cow, and \horse, though not in \sheep, \smonkey. %
%
\textit{HSD3B1}, one of two enzymes that synthesizes progesterone from
pregnalone
(\cite{Simard.ea2005:Molecularbiologyof3beta-hydroxysteroid},
\cref{fig:progesterone_evolution_tree}) is also highly expressed in
\human, \bonobo, \cow, \horse, and \dog, but is not highly expressed
in \smonkey. The diversity of \textit{CYP19A1} and \textit{HSD3B1}
expression levels in in primates and ungulates seen here points to the
need for more complete sampling of placentas at different timepoints
and in different clades.
